---
title: "Clustered Energy Consumption Patterns"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: rows
runtime: shiny
---

<!-- Learn more about flexdashboard at https://rstudio.github.io/flexdashboard -->


```{r setup, include=FALSE}
# Load packages and initialize data here
library(flexdashboard)
library(dplyr)
library(ggplot2)
library(tidyr)
library(dygraphs)
library(lubridate)
library(data.table)
library(RColorBrewer)
library(plotly)


rm(list=ls())

source('aux.R')


```

Time Series
=======================================================================

Column {.sidebar}
-----------------------------------------------------------------------
```{r inputs}
 selectInput(
        'clusters', strong('Select clusters to visualize'), choices = 1:(ncol(series)-1),
            multiple = TRUE, selected=1:5
      )

#inestable
#observeEvent(input$clusters2, {updateSelectInput(session, "clusters", selected=input$clusters2)})
```

Click and drag for zoom. Double click to reset zoom.
The number in the bottom-left corner controls the width of rolling average span: 1 displays hourly values, 24 for daily moving average and so on.

-----

```{r input_fields}
selectInput("field", strong("Classification Field"), choices=list("CNAE","MERCADO","power_class", "CUPS_class"),
            multiple=FALSE)

```

* **CNAE**: Custormer Type. Household (T1) or bussines (T2)
* **MERCADO**: Regulated Tariff (M1) or free market price (M2)
* **power_class**: Contracted capacity class (quartile split). In W.
* **CUPS_class**: TARGET_TENENCIA_CUPS class (quartile split, 0's grouped into a single category). TARGET_TENENCIA_CUPS measures the probability of Gas supply available in the customer area.

Missing values frequencies are also displayed when present (dark grey bars)

Row
-----------------------------------------------------------------------

### Time Line

```{r dy_gr}
renderDygraph({
    if (length(input$clusters)==0)
        return(NULL)
    dygraph_df(series[, c(1,as.integer(input$clusters)+1)] %>% filter(fecha>fecha_i), group='gr1') %>% 
        dyRoller(rollPeriod = 24) %>% 
        dyAxis("y", label = "Proportion of contracted power") %>%
        dyOptions(includeZero = TRUE)
})
```

Row
-----------------------------------------------------------------------

###

```{r, plot1}

renderPlot({
    compositionPlot(field=input$field)
})

```

###

```{r, plot2}
renderPlot({
    frequencyDevPlot(field=input$field)
})
```



```{r}
# Code for Chart C
```

Averaged Patterns
=======================================================================

Column {.sidebar}
-----------------------------------------------------------------------
```{r inputs2}
 selectInput(
        'clusters2', strong('Select clusters to visualize'), choices = 1:(ncol(series)-1),
            multiple = TRUE, selected=1:5
      )

#inestable
#observeEvent(input$clusters, {updateSelectInput(session, "clusters2", selected=input$clusters)})

```

Hover on the series to query associated values.

Row
-----------------------------------------------------------------------
### Week day average values
```{r gr_semana}
#renderPlot({plot(grafico_promedios(series_wday_m, 'wday', input$clusters2))})
renderPlotly({ggplotly(grafico_promedios(series_wday_m, 'wday', input$clusters2))})
```

### Week day deviation from average
```{r gr_semana_dev}
renderPlotly({ggplotly(grafico_promedios(mean_dev_wday, 'wday', input$clusters2))})
```

Row
-----------------------------------------------------------------------
### Monthly average values
```{r gr_mes}
renderPlotly({ggplotly(grafico_promedios(series_month_m, 'month', input$clusters2))})
```

### Monthly deviation from average
```{r gr_mes_dev}
renderPlotly({ggplotly(grafico_promedios(mean_dev_month, 'month', input$clusters2))})
```



Hourly Averaged Patterns
=======================================================================

Row 
-----------------------------------------------------------------------

### Hourly average consumption by month
```{r hour_month_plot}
 renderPlot({plot(gr_h_month)})
#plot(gr_h_month)
```
